Methods for understanding the variable importance of local explanations of black-box models

### 2021-10-25 _paper.rmd
abstract

---
title: "Methods for understanding the variable importance of local explanations of black-box models"
# author:
#   - name: Nicholas Spyrison
#     address: |
#       | Monash University
#       | Faculty of Information Technology
#       | ORCiD: 0000-0002-8417-0212
#     email: \email{nicholas.spyrison@monash.edu}
#     orcid: 0000-0002-8417-0212
#   - name: Dianne Cook
#     affiliation: Monash University
#     address: |
#       | ORCiD: 000-0002-3813-7155
#     orcid: 000-0002-3813-7155
#   - name: Kimbal Marriott
#     affiliation: Monash University
#     address: |
#       | ORCiD: 0000-0002-9813-0377
#     orcid: 0000-0002-9813-0377
# keywords:
#   formatted: [XAI, explanable artificial intelligence, black-box models, local explanations, grand tour, manual tour, DR, dimension reduction, R]
#   plain:     [XAI, explanable artificial intelligence, black-box models, local explanations, grand tour, manual tour, DR, dimension reduction, R]
bibliography: cheem_paper.bib
abstract: |
  ABSTRACT NEEDED
# documentclass: jdssv
# classoption: article
output:
  bookdown::pdf_document2:
    keep_tex: true
    toc: false
    #template: template/jdssv_template.tex
    number_sections: yes
    fig_caption: yes
appendix:
  # - "appendix-1.Rmd"
# title:
#   formatted: "Methods for understanding the variable importance of local explanations of black-box models"
#   plain:     "Methods for understanding the variable importance of local explanations of black-box models"
#   short:     "Use of the manual tour to visually interrogate local explanations"
editor_options:
  chunk_output_type: console
urlcolor: blue
linkcolor: red
header-includes:
 - \usepackage{setspace} ### for title page spacing
 - \usepackage{hyperref} ### for all sorts of linking
 - \usepackage{graphicx} ### to insert Monash crest
---
```{r include=FALSE, cache=FALSE}
require("knitr")
require("kableExtra")
require("magrittr")
## Work packages
require("cheem")
require("spinifex")
## chunk options
knitr::opts_chunk$set(
  fig.align = "center",
  echo = FALSE, ## Code
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  error = FALSE,
  cache = TRUE,
  cache.lazy = FALSE)
```

@@

# Abstract

Artificial Intelligence (AI) has seen a revitalization in recent years from the use of increasingly hard-to-interpret black-box models. The lack of transparency in such models has actualized as research toward explainable AI (XAI). While there are

@@

# Interrogate variable imporances of the local explanations: Cheem
## RF model
## shap matrix
## Linked global approximations of data and local explanations
## Applcation of the manual tour to

# Design of cheem
## 1) Penguins speicies classification
## 2) FIFA wage regression 
## 3)?
## 4)?

# Software Infrastructure

>>>>

---
title: "Methods for understanding the variable importance of local explanations of black-box models"
# author:
#   - name: Nicholas Spyrison
#     address: |
#       | Monash University
#       | Faculty of Information Technology
#       | ORCiD: 0000-0002-8417-0212
#     email: \email{nicholas.spyrison@monash.edu}
#     orcid: 0000-0002-8417-0212
#   - name: Dianne Cook
#     affiliation: Monash University
#     address: |
#       | ORCiD: 000-0002-3813-7155
#     orcid: 000-0002-3813-7155
#   - name: Kimbal Marriott
#     affiliation: Monash University
#     address: |
#       | ORCiD: 0000-0002-9813-0377
#     orcid: 0000-0002-9813-0377
# keywords:
#   formatted: [XAI, explanable artificial intelligence, black-box models, local explanations, grand tour, manual tour, DR, dimension reduction, R]
#   plain:     [XAI, explanable artificial intelligence, black-box models, local explanations, grand tour, manual tour, DR, dimension reduction, R]
bibliography: cheem_paper.bib
abstract: |
  Artificial Intelligence (AI) has seen a revitalization in recent years from the use of increasingly hard-to-interpret black-box models. In such models, increased predictive power comes at the cost of opaque factor analysis, which has led to the field of explainable AI (XAI). XAI attempts to shed light on these models, one such approach is the use of local explanations. A local explanation of a model give a point-estimate of linear variable importance in the vicinity of one observation. We extract explanations for each observation, and approximate data and this attribution space side-by-side with linked brushing. After identifying an observation of interest its local explanation is used as a 1D projection basis. We then manipulate the magnitude of the variable contributions with a technique called the tour. This tour animates many projections over small changes in the projection basis. Doing so allows a user to visually explore the data space through the lens of this local explanation and interrogate its variable importance. The implementation of our framework is available as an R package called cheem available at [github.com/nspyrison/cheem](https://github.com/nspyrison/cheem).
# documentclass: jdssv
# classoption: article
output:
  bookdown::pdf_document2:
    keep_tex: true
    toc: false
    #template: template/jdssv_template.tex
    number_sections: yes
    fig_caption: yes
appendix:
  # - "appendix-1.Rmd"
# title:
#   formatted: "Methods for understanding the variable importance of local explanations of black-box models"
#   plain:     "Methods for understanding the variable importance of local explanations of black-box models"
#   short:     "Use of the manual tour to visually interrogate local explanations"
editor_options:
  chunk_output_type: console
urlcolor: blue
linkcolor: red
header-includes:
 - \usepackage{setspace} ### for title page spacing
 - \usepackage{hyperref} ### for all sorts of linking
 - \usepackage{graphicx} ### to insert Monash crest
---
```{r include=FALSE, cache=FALSE}
require("knitr")
require("kableExtra")
require("magrittr")
## Work packages
require("cheem")
require("spinifex")
## chunk options
knitr::opts_chunk$set(
  fig.align = "center",
  echo = FALSE, ## Code
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  error = FALSE,
  cache = TRUE,
  cache.lazy = FALSE)
```

@@

@@

# Interrogate variable imporances of the local explanations: Cheem
## RF model
## shap matrix
## Linked global approximations of data and local attribution spaces
## Applcation of the manual tour to intterogate local explanations
## Classification task
## Regression task

# Application Design
## 1) Penguins speicies classification
## 2) FIFA wage regression
## 3)?
## 4)?

# Software Infrastructure
## Extend spinifex, consume DALEX & treeshap
## Preprocess
## Runtime rendering

### 2021-10-25 _paper.tex
abstract

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames*,x11names*}{xcolor}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Methods for understanding the variable importance of local explanations of black-box models},
  pdfkeywords={blah, blah},
  colorlinks=true,
  linkcolor=red,
  filecolor=Maroon,
  citecolor=Blue,
  urlcolor=blue,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{setspace}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\ifluatex
  \usepackage{selnolig}  % disable illegal ligatures
\fi

\title{Methods for understanding the variable importance of local explanations of black-box models}
\author{true \and true}
\date{}

\begin{document}
\maketitle
\begin{abstract}
ABSTRACT NEEDED
\end{abstract}

\hypertarget{abstract}{%
\section{Abstract}\label{abstract}}

@@

\hypertarget{linked-global-approximations-of-data-and-local-explanations}{%
\subsection{Linked global approximations of data and local explanations}\label{linked-global-approximations-of-data-and-local-explanations}}

\hypertarget{applcation-of-the-manual-tour-to}{%
\subsection{Applcation of the manual tour to}\label{applcation-of-the-manual-tour-to}}

\hypertarget{design-of-cheem}{%
\section{Design of cheem}\label{design-of-cheem}}

@@

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames*,x11names*}{xcolor}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Methods for understanding the variable importance of local explanations of black-box models},
  colorlinks=true,
  linkcolor=red,
  filecolor=Maroon,
  citecolor=Blue,
  urlcolor=blue,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{setspace}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\ifluatex
  \usepackage{selnolig}  % disable illegal ligatures
\fi

\title{Methods for understanding the variable importance of local explanations of black-box models}
\author{}
\date{\vspace{-2.5em}}

\begin{document}
\maketitle
\begin{abstract}
Artificial Intelligence (AI) has seen a revitalization in recent years from the use of increasingly hard-to-interpret black-box models. In such models, increased predictive power comes at the cost of opaque factor analysis, which has led to the field of explainable AI (XAI). XAI attempts to shed light on these models, one such approach is the use of local explanations. A local explanation of a model give a point-estimate of linear variable importance in the vicinity of one observation. We extract explanations for each observation, and approximate data and this attribution space side-by-side with linked brushing. After identifying an observation of interest its local explanation is used as a 1D projection basis. We then manipulate the magnitude of the variable contributions with a technique called the tour. This tour animates many projections over small changes in the projection basis. Doing so allows a user to visually explore the data space through the lens of this local explanation and interrogate its variable importance. The implementation of our framework is available as an R package called cheem available at \href{https://github.com/nspyrison/cheem}{github.com/nspyrison/cheem}.
\end{abstract}

@@

\hypertarget{linked-global-approximations-of-data-and-local-attribution-spaces}{%
\subsection{Linked global approximations of data and local attribution spaces}\label{linked-global-approximations-of-data-and-local-attribution-spaces}}

\hypertarget{applcation-of-the-manual-tour-to-intterogate-local-explanations}{%
\subsection{Applcation of the manual tour to intterogate local explanations}\label{applcation-of-the-manual-tour-to-intterogate-local-explanations}}

\hypertarget{classification-task}{%
\subsection{Classification task}\label{classification-task}}

\hypertarget{regression-task}{%
\subsection{Regression task}\label{regression-task}}

\hypertarget{application-design}{%
\section{Application Design}\label{application-design}}

@@

\hypertarget{extend-spinifex-consume-dalex-treeshap}{%
\subsection{Extend spinifex, consume DALEX \& treeshap}\label{extend-spinifex-consume-dalex-treeshap}}

\hypertarget{preprocess}{%
\subsection{Preprocess}\label{preprocess}}

\hypertarget{runtime-rendering}{%
\subsection{Runtime rendering}\label{runtime-rendering}}
